### 2014-05-14 doc/ApTranscriptome_manuscript.Rmd
Small edits

Colonies of the ant species *Aphaenogaster picea* were collected from Molly Bog, Vermont (University of Vermont Natural Areas; 44.508°N, -72.702°W) and *Aphaenogaster carolinensis* from Durham, North Carolina ( 36.037° N -78.874° W), designated as *ApVT* and *AcNC* respectively. The phylogeny of *Aphaeogaster* is currently unresolved [@umphrey1996], but putative species identification was possible based on morphology (Bernice Bacon DeMarco, pers. communication). species of these species were maintained in the lab for 6 months prior to experimentation.

From each species, we haphazardly collected 12 ants at the same time on 12 days. Each day, the 12 ants were placed in glass tubes into a water bath at one of 12 randomly-assigned temperatures, every 3.5°C between 0° and 38.5°C, for one hour. The minimum and maximum temperatures were selected based on preliminary work showing that these temperatures are close to the critical minimum and maximum temperatures for *Aphaenogaster* [@warren2013]. At the end of the hour, the ants were flash frozen in liquid nitrogen and stored at -80°C. We extracted mRNA by homogenizing three pooled ants in 500 uL of RNAzol buffer with with zirconium silicate beads in a Bullet Blender (Next Advance; Averill Park, NY), followed by RNAzol extraction (Molecular Research Center Inc; Cincinnati, OH) and then a RNeasy micro extraction (Qiagen Inc; Valencia, CA) following the manufacturer's instructions.

@@

For each species, the 12 samples were barcoded and sequenced in a single lane of 2 x 100bp paired-end reads on an Illumina HiSeq 1500 yielding 200 and 160 million reads for the A. picea and A. carolinensis samples respectively. Reads were filtered to remove Illumina adapter sequences and low quality bases using the program Trimmomatic [@lohse2012].

We assembled the sequenced reads into the full set of mRNA transcripts (the `transcriptome`) for the combined data set from both species using the Trinity *de novo* transcriptome assembly program [@grabherr2011]. *De novo* transcriptome assembly is prone to falsely identifying alternative transcripts and identifying inaccurate transcripts that are chimeric (e.g. regions of two separate transcripts with enough overlap to assemble into a false, or chimeric, third transcript) [@yang2013] To account for this, we first ran program `CAP3` [@huang1999] to cluster sequences with greater than 90% similarity and merge transcripts with overlaps longer than 100 bp and 98% similar in length, and second ran the program `uclust` which cluster sequences completely contained within longer sequences at greater than 90% similarity (see Supplemental for exact parameters). We used liberal values (90% similarity) to merge transcripts to account for the fact that we were including sequencing data from two species of the same species, and wanted to avoid assembling orthologs individually. We performed subsequent analyses at the transcript level due to limitations of working with a system lacking a reference genome, but refer to these transcripts as genes for simplicity.

We performed functional annotation of the transcriptome assembly using the web-based tool `FastAnnotator` [@chen2012] which annotates and classifies genes by Gene Ontology (GO) term assignment, enzyme identification and domain identification.

@@

We quantified expression of each gene using the program `Sailfish` [@patro2013]. `Sailfish` reports gene expression as transcripts per million (TPM) which is the preferred measure of gene expression as values are comparable among sequencing runs [@wagner2012]. In addition, `Sailfish` corrects for sequence composition bias and gene length. As preliminary examination of the data (see supplemental...) indicated that 7°C samples may have been mis-labeled, we omitted these data from the analysis. The expression values were highly correlated between species at each temperature treatment (r^2^ > 0.98) indicating that assembling the transcriptome with data from both species was justified (Supplemental).

@@

To gain greater insight on the genetic basis of thermal tolerance in *Aphaenogaster*, we performed comparative analyses between the two species focusing on the thermally-responsive genes that had a significant species x temperature interaction. For each gene, we predicted expression levels across temperatures for each species using the full linear model, and then grouped genes into the four responsive categories, *High*, *Low*, *Intermediate*, and *Bimodal*, as well as a fifth category *Not Expressed*. To focus on differences in gene expression between the two species, we performed GSEA for the set difference of each category.

@@

The Trinity *de novo* transcriptome assembly included 126,172 genes with a total length of 100 million bp. Filtering to remove redundant or chimeric reads resulted in an assembly with 96,253 contigs and a total of 105,536 genes. For all genes, the total length was 63 million bp with an N~50~ length of 895 bp and a mean contig size of 593 bp. Of the 105,536 filtered genes, 55,432 had hits to the NCBI-nr database. Of these, 38,711 genes mapped to GO terms, 1,659 genes were identified to an enzyme and 18,935 genes mapped to a domain with >50% coverage. Of these, 5,787 genes have best hits to putative genes in the genome sequence of the ant species *Solenopsis invicta* and XXXX are annotated to genes previously identified from organisms within the class Insecta.

>>>>

Colonies of the ant species *Aphaenogaster picea* were collected from Molly Bog, Vermont (University of Vermont Natural Areas; 44.508°N, -72.702°W) and *Aphaenogaster carolinensis* from Durham, North Carolina ( 36.037° N -78.874° W), designated as *ApVT* and *AcNC* respectively. The phylogeny of *Aphaeogaster* is currently unresolved [@umphrey1996], but putative species identification was possible based on morphology (Bernice Bacon DeMarco, pers. communication). Colonies of these species were maintained in the lab for 6 months prior to experimentation.

From each species, we haphazardly collected 12 ants at the same time on 12 days. Each day, the 12 ants were placed in glass tubes into a water bath at one of 12 randomly-assigned temperatures, every 3.5°C between 0° and 38.5°C, for one hour. The minimum and maximum temperatures were selected based on previous work showing that these temperatures are close to the critical minimum and maximum temperatures for *Aphaenogaster* [@warren2013]. At the end of the hour, the ants were flash frozen in liquid nitrogen and stored at -80°C. We extracted mRNA by homogenizing three pooled ants in 500 uL of RNAzol buffer with zirconium silicate beads in a Bullet Blender (Next Advance; Averill Park, NY), followed by RNAzol extraction (Molecular Research Center Inc; Cincinnati, OH) and then an RNeasy micro extraction (Qiagen Inc; Valencia, CA) following the manufacturer's instructions.

@@

For each species, the 12 samples were barcoded and sequenced in a single lane of 2 x 100bp paired-end reads on an Illumina HiSeq 1500 yielding 200 and 160 million reads for the *A. picea* and *A. carolinensis* samples respectively. Reads were filtered to remove Illumina adapter sequences and low quality bases using the program Trimmomatic [@lohse2012].

We assembled the sequenced reads into the full set of mRNA transcripts, the transcriptome, for the combined data set from both species using the Trinity *de novo* transcriptome assembly program [@grabherr2011]. *De novo* transcriptome assembly is prone to falsely identifying alternative transcripts and identifying inaccurate transcripts that are chimeric (e.g. regions of two separate transcripts that assemble into a false, or chimeric, third transcript) [@yang2013]. We removed potentially false transcripts by first running the program `CAP3` [@huang1999] to cluster sequences with greater than 90% similarity and merge transcripts with overlaps longer than 100 bp and 98% similar in length. Second, we ran the program `uclust` which clusters sequences completely contained within longer sequences at greater than 90% similarity (see Supplemental Information for exact parameters). We used liberal values (90% similarity) to merge orthologous transcripts in the two species that may not have assembled together in the initial *de novo* transcriptome assembly. We performed subsequent analyses at the transcript level due to limitations of working with a system lacking a reference genome, but refer to these transcripts as genes for simplicity.

To determine the putative function of the transcripts, we performed functional annotation of the transcriptome assembly using the web-based tool `FastAnnotator` [@chen2012] which annotates and classifies genes by Gene Ontology (GO) term assignment, enzyme identification and domain identification.

@@

We quantified expression of each gene using the program `Sailfish` [@patro2013] and use the bias-corrected transcripts per million (TPM) [@wagner2012] as our measure of gene expression. As preliminary examination of the data (see supplemental...) indicated that 7°C samples may have been mis-labeled, we omitted these data from the analysis. The expression values were highly correlated between species at each temperature treatment (r^2^ > 0.98) indicating that assembling the transcriptome with data from both species was justified (Supplemental).

@@

To gain greater insight on the genetic basis of thermal tolerance in *Aphaenogaster*, we performed comparative analyses between the two species focusing on the thermally-responsive genes that had a significant species x temperature interaction. For each gene, we predicted expression levels across temperatures for each species using the full linear model, and then grouped genes into the four responsive categories, *High*, *Low*, *Intermediate*, and *Bimodal*, as well as a fifth category *Not Responsive*. To focus on differences in gene expression between the two species, we performed GSEA for the set difference of each category.

@@

The Trinity *de novo* transcriptome assembly included 126,172 genes with a total length of 100 million bp. Filtering to remove redundant or chimeric reads resulted in an assembly with 96,253 contigs and a total of 105,536 genes. For all genes, the total length was 63 million bp with an N~50~ length of 895 bp and a mean contig size of 593 bp. Of the 105,536 filtered genes, 55,432 had hits to the NCBI-nr database. Of these, 38,711 genes mapped to GO terms, 1,659 genes were identified to an enzyme and 18,935 genes mapped to a domain with >50% coverage. Of these, 5,787 genes have best hits to putative genes in the genome sequence of the ant species *Solenopsis invicta* and 47% have best hits to genes from the class Insecta.
